Report on Late Winter2013 Production: Image Differencing

### 2013-03-26 W13report.tex
Good AJC point wanted to capture.
Added some discussion of WCS failure modes and measurement on negative sources.  Added figures.

% latex W13report.tex; bibtex W13report; latex W13report; latex W13report; dvips W13report; ps2pdf W13report.ps ; gv W13report.pdf

@@

We assume that science image $S(x,y)$ can be modeled as a convolution
of the template image $T(x,y)$ by a PSF--matching kernel $K(u,v,x,y)$.
We further assume that the kernel may be decomposed using a set of
basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the coefficients
in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$.  We also assume that
there is a differential background between the two images $b(x,y)$
that may be fit for using a low--order Chebyshev.  The image
difference is then calculated through $D(x,y) = S(x,y) - T(x,y)
\otimes K(u,v,x,y) - b(x,y)$.

@@

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we detect on the difference image directly.
In post--filtering, we correlate the difference image with its Psf (by
design, the same as the Psf of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  However, differences in the
shapes and sizes of the Psf--matching kernels used in the two cases
may cause systematic differences in the images.  Detections are merged
using a grow radius of 2 pixels.

@@

A template image was generated by {\bf DETAIL}.  The template has a
seeing of 0.88''.  This means that the matching of the template image
with visit 6866601 will yield a template {\it deconvolution} in the
postfilter configuration.  Matching of the template image with visit
6866602 also yields a mild deconvolution (the images were generated to
have the same seeing) for 5/9 sensors.  The matching of the template
with visit 6866603 with postfiltering, and of all visits with
prefiltering, yields a convolution (smoothing) of the image.

@@

@@

Because these images will have gone through multiple convolutions
before reaching the detection stage, it is expected that the
propagated per-pixel variance will not provide an exact representation
of the empirical variance in the image planes.  Because the
astrometric resampling uses a warping kernel that is designed to
preserve the noise properties of the images, we expect that the kernel
convolution and Psf--filtering will be the largest sources of error.

To quantify these errors, we first compared the empirical variance in
the image planes with the level tracked by the variance planes.  We
calculated the ratio of the interquartile range of unmasked pixels in
the image plane, multiplied by 0.741 and then squared, with the median
of unmasked pixels from the variance plane.  The former represents the
empirical variance, while the latter the propagated variance.  We
investigate their ratio (empirical/propagated) at 2 stages in the
processing: after creation of the difference image (image $D$), and
immediately preceding detection (image $D'$).  Recall that for
prefiltering, $D = D'$.  These results are summarized in
Table~\ref{tab-variance1}.  For prefiltering, we find that the
variance plane typically underestimates the true variance by $1\%$ for
all visits.  When using postfiltering, the deconvolution visit
v6866601 yields a similar underestimate in the difference image, but
for the other two visits the variance is tracked to better than 1\%,
with a small RMS.  However, when postfiltering with the Psf, the
variance becomes {\it overestimated} by nearly a factor of two for
deconvolution visit v6866601, and {\it underestimated} by $4-5\%$ for
the other visits.

We further investigate the variance properties of the difference
images immediately preceding detection, i.e. the difference image
itself in the case of prefiltering, and the difference image
correlated with its Psf in postfiltering.  These results are
summarized in Table~\ref{tab-variance2}.  We find that, empirically,
the images have very similar noise properties at the detection stage,
with the deconvolution visit having larger variance by $1\%$ but the
other visits having essentially equal noise properties, with a bias
towards the postfiltered data having slightly higher variance (while
the RMS is small, all values are greater than 1.0).  The medians of
the variance planes tell a different story though.  For the
deconvolution visit, the median per--pixel variance is $70\%$ higher.
While this will suppress the detections of false positives, it will
also significantly suppress the detection efficiency for true
variability.  For the other visits, the prefiltered per--pixel
variance is relatively {\it higher} by $3-4\%$.  This will decrease
the sensitivity of the prefiltered data to false positives; as shown
above, the rate is indeed smaller using the prefiltering pipeline.
This analysis demonstrates that this is primarily due to an {\it
  underestimate} of the per--pixel variance in the postfilter
processing.  As outlined by \cite{Price-Stacking}, this may be
accounted for by differences in pixel covariance when undergoing
multiple convolutions, which are not currently tracked.  We have not
yet definitively demonstrated that this is the case here.

@@

@@

\clearpage

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{4}{|c|}{Variance Ratios: Empirical / Propagated} \\
\hline
Visit    & $D_{pre} = D'_{pre}$ & $D_{post}$ & $D'_{post}$ \\
\hline
v6866601 &$1.012 \pm 0.002$&$1.018 \pm 0.005$&$0.599 \pm 0.017$ \\
v6866602 &$1.010 \pm 0.002$&$1.007 \pm 0.001$&$1.046 \pm 0.002$ \\
v6866603 &$1.015 \pm 0.003$&$1.008 \pm 0.001$&$1.045 \pm 0.003$ \\
\end{tabular}
\caption{Ratio of the empirical variance in the difference images,
  calculated via (0.741 times the interquartile range)$^2$, to the
  median of the variance plane.  In all cases the variance plane
  represents an {\it underestimate} of the true variance in the
  images.  We report the mean and RMS across all sensors for the best
  configurations.  }
\label{tab-variance1}
\end{table}

\begin{table}
\centering
\begin{tabular}{ccc}
\hline
\multicolumn{3}{|c|}{Variance Ratios: $D'_{post} / D'_{pre}$} \\
\hline
Visit    & Empirical & Propagated \\
\hline
v6866601 & $1.011 \pm 0.016$    & $1.710 \pm 0.049$    \\
v6866602 & $1.001 \pm 0.001$    & $0.967 \pm 0.001$    \\
v6866603 & $1.002 \pm 0.001$    & $0.974 \pm 0.002$    \\
\end{tabular}
\caption{Ratios of the variance planes immediately before detection
  $D'$.  Ratios are in the sense of the variance of the postfiltered
  image divided by the variance of the prefiltered image.  The first
  column represents the ratios of the empirical variance, calculated
  via (0.741 times the interquartile range)$^2$.  The second column
  reports the ratios of the means of the propagated variance planes.
  We report the mean and RMS across all sensors for the best
  configurations.}
\label{tab-variance2}
\end{table}

\clearpage

>>>>

% latex W13report.tex; bibtex W13report; latex W13report; latex W13report; dvips W13report; gv W13report.ps

@@

We assume that science image $S(x,y)$ can be modeled as a convolution
of the template image $T(x,y)$ by a PSF--matching kernel $K(u,v,x,y)$.
We further assume that the kernel may be decomposed using a set of
basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the coefficients
in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$.  We also assume that
there is a differnetial background between the two images $b(x,y)$
that may be fit for using a low--order Chebyshev.  The image
difference is then calculated through $D(x,y) = S(x,y) - T(x,y)
\otimes K(u,v,x,y) - b(x,y)$.

@@

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we detect on the difference image directly.
In post--filtering, we correlate the difference image with its Psf (by
design, the same as the Psf of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  We expect minor differences
in the variance due to the fact that the post--filtered data will have
gone through 3 convolutions (astrometric resampling; Psf--matching to
the science image; Psf filtering for detection) while the
pre--filtered data will have gone through only two (astrometric
resampling; psf--matching to the maximum likelihood image).
Differences in the shapes and sizes of the Psf--matching kernels used
in the two cases may also cause systematic differences in the images.
Detections are merged using a grow radius of 2 pixels.

@@

A template image was generated by {\bf DETAIL}.  The template has a
seeing of 0.88''.  This means that the matching of the template image
with visit 6866601 will yield a template {\it deconvolution} in the
postfilter configuration.  Matching of the template image with visit
6866602 also yields a mild deconvolution (the images were generated to
have the same seeing) for 5/9 sensors.  The matching of the template
with visit 6866603 with postfiltering, and of all visits with
prefiltereing, yields a convolution (smoothing) of the image.

@@

Section \ref{sec-analyticfp} discusses the theoretical foundation 
that sets the floor on the number of false positives we expect from a 
pure Gaussian random field.  The number of false positives is a strong
function of the seeing and and the confidence level of the detection.  
By comparing the emperical false positive rate to the analytic predictions
we can quantify how close we are getting to the theoretical floor in these
data.  Because the number of false positives is such a strong function of
detection threshold, misestimation of the signal to noise of the detections
has a large impact on how many false positives are observed.  The Figures
\ref{fig-fpthresh1}, \ref{fig-fpthresh2}, and \ref{fig-fpthresh3} show the 
effect of misestimating the noise in the detections for each of three different
seeing values (0.6\", 0.88\", 1.2\" respectively).

@@

The variance in the prefiltered and postfiltered images is likely to
be different due to the different types of operations that have
happened on the images.  We quantify this by taking the postfiltered
images, immediately before the detection step (i.e. $D'$), and compare
the variance plane to that coming from the prefiltered image
(Table~\ref{tab-variance2}).  For v6866601, which is a deconvolution
in the postfiltered production, we find that the variance is 70\%
higher when postfiltering.  This is consistent with the expectation
that deconvolution increases the noise properties of the image.  For
v6866602 and v6866603, we find that the variance is 3--4\% and 2--3\%
higher in the {\it prefiltered} data, respectively.  The primary
culprit for this is that the prefiltered data will have gone through 1
less convolution, meaning less of the variance will have been lost to
covariance.

To quantify the significance of these offsets, we compared the
empirical variance in the image planes with the variance tracked by
the variance planes to ascertain how much we are misestimating the
true variance.  We calculated the ratio of the interquartile range of
unmasked pixels in the image plane, multiplied by 0.741 and then
squared, with the median of unmasked pixels from the variance plane.
The results are summarized in Table~\ref{tab-variance1}.  We find that
we are marginally, but consistently, underestimating the variance in
our images by $\sim 1\%$.  This affect is smaller than the amplitudes
reported above, indicating that there are true differences in the
variances of prefiltered and postfiltered detection images.  As the
variance is higher in the prefiltered images, we will be slightly less
sensitive to true variability.

@@

\section{Outstanding Issues}
We ran into several issues while doing this production that required work 
arounds.  In some cases we did not have the time to drill all the way down
to the root cause.  We itemize these issues here so that they may be addressed 
as part of the Summer 2013 work.

\subsection{Astrometric Registration}
The original plan for this production was to use single deep exposures as 
templates.  This required that we treat the single exposures as coadds.  As 
such we warped the exposures to coadd patches for use in the image difference
task.  The register task is unable to solve a differential WCS if one of the images
has the CRPIX off the image.  This is the case for coadd patches.  To get around this
we resolved the coadd patches using a TAN-SIP solution and then used the register
task to solve for the differential WCS.  This worked fairly will part of the time 
yeilding residuals in the 40-100 mas range.  The rest of the time the 
residual was more like 250 mas range.  This is an order of magnitude larger than expected.

We investigated potential causes of the large residuals.  To do this, we looked at the 
residuals for template solutions, science image solutions, and template to science calexp 
registration to attempt to isolate where the astrometry was being degraded.  The results are shown
in table \ref{tab-wcsrms}.  We show the results for the nine sensors in the best seeing visit.

The solutions for the three steps that do not involve any warping show results consistent with 
the accuracy we expect to get.  We also see the expected trend that the deep exposures have 
smaller RMS than the solutions of the science images and the differential solution is dominated
by the RMS in the science image.  In four cases, the coadd to science solution did not fail
as catastrophically as in the other five, but the RMS is still a factor of several larger
than in the non-warped cases.

\subsection{Detection and Measurement on Negative Sources}
We see more negative false positives than positive false positives.  This drove an investigation 
into the symmetry of detection on positive/negative sources.  To do this we looked at the results 
of the image difference task and compared it to the same runs when the difference image was inverted
$differenceImage *= -1$ before measurement.  The results are as follows:
\begin{itemize}
\item The DIA source lists are always the same length.
\item Negative sources always have integer centroid pixel values where positive sources have fractional
pixel precision.
\item Some sources are have the edge bit set in one image or the other.  This leads to a difference in 
the number of false positives as we only consider those without the {\tt EDGE} flag set.
\end{itemize}

The suspicion is that the difference in precision of the centroid is linked to the difference in flags.
We have not had the time to chase this down completely, but plan on doing so in Summer 2013.

@@

\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Variance Ratios} \\
\hline
Visit    & Prefilter & Mean Ratio & RMS Ratio \\
\hline
v6866601 & True      & 1.019      & 0.005    \\
         & False     & 1.011      & 0.002    \\
v6866602 & True      & 1.007      & 0.001    \\
         & False     & 1.010      & 0.002    \\
v6866603 & True      & 1.008      & 0.001    \\
         & False     & 1.015      & 0.003    \\
\end{tabular}
\caption{Ratio of the empirical variance in the difference images,
  calculated via (0.741 times the interquartile range)$^2$, to the
  median of the variance plane.  In all cases the variance plane
  represents an {\it underestimate} of the true variance in the
  images.  We report the mean and RMS across all sensors.  }
\label{tab-variance1}
\end{table}

\begin{table}
\centering
\begin{tabular}{ccc}
\hline
\multicolumn{3}{|c|}{Variance Ratios} \\
\hline
Visit    & Mean Postfilter/Prefilter & RMS Postfilter/Prefilter \\
\hline
v6866601 & 1.72      & 0.04     \\
v6866602 & 0.967     & 0.001    \\
v6866603 & 0.975     & 0.002    \\
\end{tabular}
\caption{Ratios of the variance planes immediately before detection
  $D'$.  Ratios are in the sense of the variance of the postfiltered
  image divided by the variance of the prefiltered image.  We report
  the mean and RMS across all sensors.}
\label{tab-variance2}
\end{table}

\clearpage
\begin{table}
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\multicolumn{5}{|c|}{WCS RMS in mas at various stages} \\
\hline
Sensor    & Deep & Science &  Deep to Science & Coadd to Science \\
\hline
R22:S00&21&29&19&382\\ 
R22:S10&13&22&22&389\\
R22:S20&7&18&21&375\\
R22:S01&13&25&21&86\\
R22:S11&6&17&20&81\\
R22:S21&8&19&20&382\\
R22:S02&8&17&20&178\\
R22:S12&9&20&20&78\\
R22:S22&7&21&21&389\\
\hline
\end{tabular}
\caption{We looked at the resulting RMS from WCS fitting on the deep exposure alone, the science exposure alone,
fitting the deep exposure to the science exposure, and fitting a warped coadd to the science exposure.  Fitting 
the coadd to the science exposure produces an order of magnitude more scatter in the fit than at any other point.\label{tab-wcsrms}}
\end{table}
\clearpage
